Blum axioms and nondeterministic computation of functions

### 2016-04-12 paper.tex
paper.tex: Rewrite the examples of Blum complexity measures
I removed the independence of the axioms as examples
because this sounded too technical for such a short paper.

\begin{example}
    Any encoding of deterministic Turing machines as a binary string
    yields an acceptable numbering of the recursive functions.
    That is, fix one of these encodings;
    define $\pi: \{0, 1\}^* \to \mathcal P$ as follows:
    if the string $w$ encodes the Turing machine $M$,
    $\pi_w$ is the partial function computed by $M$.
\end{example}

@@

\begin{example}
    Given the Gödel numbering $\pi$ of the previous chapter,
    the number of steps a Turing machine needs before halting
    is a complexity measure that satisfies Blum axioms.
    The corresponding function $\Phi$ is
    \begin{equation*}
        \Phi(w, x) = \begin{cases}
            k, &\text{if $w$ takes exactly $k$ steps on $x$ before halting} \\
            \text{undefined}, &\text{if $w$ never halts on $x$}
        \end{cases}
    \end{equation*}
    Then, $\Phi(w, x)$ is defined exactly when $\pi_w(x)$ exists
    (that is, when $w$ halts on $x$),
    and we can determine whether $\Phi(w, x) = k$ for a given $k$
    by simulating $w$ on $x$ for exactly $k$ steps.
\end{example}

\begin{example}
    Space complexity requires some care,
    because a machine may never halt even though it remains confined
    in a limited number of cells.
    Thus, we must leave the space complexity of a machine that never halts undefined.
    Formally, define the function $\Phi$ by
    \begin{equation*}
        \Phi(w, x) = \begin{cases}
            k, &\text{if $w$ uses exactly $k$ tape cells on $x$ before halting} \\
            \text{undefined}, &\text{if $w$ never halts on $x$}
        \end{cases}
    \end{equation*}
    This makes $\Phi$ satisfy the first axiom.
    To satisfy the second axiom,
    we must note that,
    if we limit the number of tape cells to $k$,
    there is a finite number of different computation states
    the machine can be.
    Therefore,
    to determine whether $\Phi(w, x) = k$,
    simulate $w$ on $x$ and keep the whole history of computation.
    If, in any point of the simulation,
    the machine $w$ uses more than $k$ tape cells,
    reject --- the answer is ``no''.
    If any two computation states repeat,
    then the machine will loop and never halt
    --- the answer is ``no'' again.
    Otherwise, the machine must halt
    (since there is a finite number of computation states).
    Now, when the machine halts,
    we just need to make sure it actually used exactly $k$ cells in one of its tapes.
\end{example}

\begin{example}
    Choosing $\Phi(w, x) = 0$ for all $w$ and $x$
    creates a function that satisfies the second axiom,
    but not the first.
    Leaving $\Phi(w, x)$ undefined for all $w$ and $x$ such that
    $\pi_w(x)$ does not exists and defining $\Phi(w, x) = 0$ otherwise
    satisfies the first axiom, but not the second.
    This shows the axioms are independent \cite[p.~324]{Blum1967}.
\end{example}

There are other measures that satisfies the Blum axioms
(like, for example,
the number of times a Turing machine changes a symbol in the tape,
the number of times the tape head changes direction,
the number of movements to the left or to the right),
but, before extending these complexity measures to nondeterministic machines,
we must first provide an acceptable Gödel numbering.

>>>>

\begin{example}\label{standard-numbering}
    Any encoding of deterministic Turing machines as a binary string
    yields an acceptable numbering of the recursive functions.
\end{example}

@@

\begin{example}
    The standard measures of time and space can be constructed
    over the acceptable numbering of example~\ref{standard-numbering}.
    The time complexity is simply the number of moves the machine makes before halting;
    if it does not halt, leave it undefined.
    As this number is only defined if the machine halts,
    the first axiom is satisfied;
    and, to decide the predicate of the second axiom,
    we can simply run the machine for $k$ moves and see whether it halted.

    Space complexity requires extra care,
    because the machine might not halt
    even though the number of tape cells it uses is limited.
    So, define the corresponding function $\Phi$ by
    \begin{equation*}
        \Phi(w, x) = \begin{cases}
            k, &\text{if $w$ scans exactly $k$ tape cells and halts when processing $x$}\\
            \text{undefined}, &\text{if $w$ never halts on $x$}
        \end{cases}
    \end{equation*}
    This function satisfies the first axiom,
    and, as there is a finite number of states of computation
    in which at most $k$ tape cells are scanned,
    the predicate of the second axiom can be verified
    by keeping all intermediate steps of computation
    and checking to see whether any state repeat
    (in this case, the machine loops and the complexity is undefined).
\end{example}

\begin{example}
    As with acceptable Gödel numberings,
    we can also measure less usual resources.
    For example, we can satisfy the Blum axioms by
    counting the number of times a machine changes a symbol in the tape,
    the number of times the tape head moves right,
    or the number of times a two-stack pushdown automata
    pops a symbol from the second stack.
\end{example}

Before extending these complexity measures to nondeterministic machines,
we must first provide an acceptable Gödel numbering.
